---
title: "Brain deviations in schizophrenia cohort"
author: "Anna Huang"
date: '2023-11-16'
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    code_folding: hide 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '/Users/ashuang/Library/CloudStorage/OneDrive-VUMC/Work/05_OngoingProjects/Lifespan')
#knitr::opts_knit$set(root.dir = 'C:/Users/ashua/OneDrive - VUMC/Work/05_OngoingProjects/Lifespan/')
```

```{r, echo=FALSE}
library(table1)
library(magrittr)
library(dplyr)
library(ggplot2)
library(reshape2)
library(patchwork)
library(scales)
library(rstatix)
library(confintr)
```

# Summary

In this project we are examining the centile results from GAMLSS models.

# Data

Loading in data about centiles so that we can do group comparions in the schizophrenia cohort.

```{r}
data = read.csv("KaidiReports/data_centiles_Oct20.csv")
behavdata = read.csv("Data_Management/PSYXSEC/PSYXSECPhenoDataSCIPPANSS.csv")
hcpbehav = read.csv("Data_Management/HCP_cognitive.csv")
#data = read.csv("D:/OneDrive - VUMC/Work/Lifespan/KaidiReports/data_centiles_Sept27.csv")
```


```{r}
# Subset data into the HCP dataset and PSYXSEC datasets
data$dx_group = data$Diagnosis
data$dx_group[data$Diagnosis != "Normal Control"] = "Schizophrenia Spectrum"
data$dx_group <- as.factor(data$dx_group)
hcp <- subset(data,Dataset == "HCP")
psyxsec <- subset(data,Dataset == "PSYXSEC")
Nucleus <- c("AV","AV","VA","VA","VL","VL","VPL","VPL",
                  "Pul","Pul","LGN","LGN","MGN","MGN","CM","CM","MD","MD")
Hemisphere <- c("Right","Left","Right","Left","Right","Left","Right","Left",
                  "Right","Left","Right","Left","Right","Left","Right","Left","Right","Left")
```

# Comparing centile medians between groups

Present the centile values in a table that includes the median centile values for SZ and HC, then compare group medians using the Wilcoxon Rank Sum test

```{r}
#### Right AV
## plot SZ vs HC Centiles cores as density maps
w.test <- data.frame(matrix(data=NA,nrow=18,ncol=9))
colnames(w.test) <- c("Nucleus","Hemisphere","SZmed","HCmed","W","L.conf","H.conf","shift","pval")
w.test$Nucleus <- Nucleus
w.test$Hemisphere <- Hemisphere

for (i in 1:18) {
  variablename <- paste('centile_',substr(w.test$Hemisphere[i],1,1),'.',w.test$Nucleus[i],sep="")
  SZ.med <- median(psyxsec[psyxsec$dx_group=="Schizophrenia Spectrum",variablename])
  HC.med <- median(psyxsec[psyxsec$dx_group=="Normal Control",variablename])
  test.results <- wilcox.test(psyxsec[,variablename]~psyxsec[,"dx_group"],conf.int=TRUE)
  w.test[i,3] <- SZ.med
  w.test[i,4] <- HC.med
  w.test[i,"W"] <- test.results$statistic
  w.test[i,"pval"] <- test.results$p.value
  w.test[i,"L.conf"] <- test.results$conf.int[1]
  w.test[i,"H.conf"] <- test.results$conf.int[2]
  w.test[i,"shift"] <- test.results$estimate
}

w.test$fdr <- p.adjust(w.test[,"pval"], method="fdr")
print(w.test)
```

# Calculating deviance status

Identifying infra-normal and supra-normal individuals in the schizophrenia cohort.

Infra-normal = lower than 5% of population
Supra-normal = greater than 95% of population

```{r}
for (i in 1:18) {
    infra <- as.numeric(psyxsec[,paste('centile_',substr(Hemisphere[i],1,1),'.',Nucleus[i], sep='')] < 0.05)
    supra <- as.numeric(psyxsec[,paste('centile_',substr(Hemisphere[i],1,1),'.',Nucleus[i], sep='')] > 0.95)
    centcat <- (infra*-1) + supra
    psyxsec <- cbind(psyxsec,centcat)
    colnames(psyxsec)[ncol(psyxsec)] <-
      c(paste(Nucleus[i],'_',substr(Hemisphere[i],1,1),'.',"deviance",sep=""))
  }

#write.csv(psyxsec,"psyxsec.csv")
```

# Identify HC individuals falling within 90% of population

Checking validity of normative models when applied to schizophrenia dataset. Here we identify the proportion of healthy individuals that fall within the middle 90% of the population. For a valid model, proportion should fall around .9.

```{r}
prop.hc <- data.frame(matrix(data=NA,nrow=18,ncol=4))
colnames(prop.hc) <- c("Nucleus","Hemisphere","Schizophrenia","Healthy Controls")
prop.hc$Nucleus <- Nucleus
prop.hc$Hemisphere <- Hemisphere
group.names <- c("Schizophrenia Spectrum","Normal Control")


for (i in 1:18) {
#  variablename <- paste('centile_',substr(w.test$Hemisphere[i],1,1),'.',w.test$Nucleus[i],sep="")
  variablename <- paste(Nucleus[i],'_',substr(Hemisphere[i],1,1),'.',"deviance",sep="")
  prop.hc[i,3] <- (sum(psyxsec[psyxsec$dx_group=="Schizophrenia Spectrum",variablename]==0)/234) * 100
  prop.hc[i,4] <- (sum(psyxsec[psyxsec$dx_group=="Normal Control",variablename]==0)/211) * 100
}

print(prop.hc)
```


# Identifying proportion of infra- and supra-normal individuals in schizophrenia cohort

Identify the proportions of individuals that fall outside 90% of the population, and then plot the proportions of infra-normal and supra-normal individuals in the healthy control and schizophrenia groups (Figure 3).

```{r}
prop.90 <- data.frame(matrix(data=NA,nrow=18,ncol=6))
colnames(prop.90) <- c("Nucleus","Hemisphere","SZinfra","SZsupra","HCinfra","HCsupra")
prop.90$Nucleus <- Nucleus
prop.90$Hemisphere <- Hemisphere
group.names <- c("Schizophrenia Spectrum","Normal Control")

for (i in 1:18) {
  variablename <- paste(Nucleus[i],'_',substr(Hemisphere[i],1,1),'.',"deviance",sep="")
  prop.90[i,3] <- (sum(psyxsec[psyxsec$dx_group=="Schizophrenia Spectrum",variablename]==-1)/234) * 100
  prop.90[i,4] <- (sum(psyxsec[psyxsec$dx_group=="Schizophrenia Spectrum",variablename]==1)/234) * 100
  prop.90[i,5] <- (sum(psyxsec[psyxsec$dx_group=="Normal Control",variablename]==-1)/211) * 100
  prop.90[i,6] <- (sum(psyxsec[psyxsec$dx_group=="Normal Control",variablename]==1)/211) * 100
}

print(prop.90)

infra.long <- melt(prop.90, id = c("Nucleus","Hemisphere"), 
                  measure = c("SZinfra","HCinfra"), 
                  variable.name = "Group",
                  value.name = "Infra")

infra.long$Group <- gsub("SZinfra","SZ",infra.long$Group)
infra.long$Group <- gsub("HCinfra","HC",infra.long$Group)

supra.long <- melt(prop.90, id = c("Nucleus","Hemisphere"), 
                  measure = c("SZsupra","HCsupra"), 
                  variable.name = "Group",
                  value.name = "Supra")

supra.long$Group <- gsub("SZsupra","SZ",supra.long$Group)
supra.long$Group <- gsub("HCsupra","HC",supra.long$Group)

all.long <- merge(x = infra.long, y = supra.long, by.x = c("Nucleus","Hemisphere","Group"), by.y = c("Nucleus","Hemisphere","Group"))

rm(infra.long,supra.long)
```

```{r  fig1,dev='png',dpi=300}

p5.inf <- ggplot(data=all.long, aes(x=Nucleus,y=Infra,fill=Group)) + facet_wrap(~ Hemisphere)
p5.sup <- ggplot(data=all.long, aes(x=Nucleus,y=Supra,fill=Group)) + facet_wrap(~ Hemisphere)



p5.infplot <- p5.inf + geom_bar(stat="identity",position=position_dodge()) + ggtitle('Infra-normal') +
  geom_hline(yintercept=5, linetype="dashed", color = "black") +
 theme(axis.text.x = element_text(angle = 90, size=14,face="bold"), 
        text = element_text(size = 14,face="bold"),
        axis.line = element_line(colour="black"),
        panel.background = element_blank(),
        plot.title = element_text(size = 16, face = "bold")) +
  labs(y="Percentage (%)",x="") + 
  scale_fill_manual(breaks=c("HC","SZ"),values=c("#00BFC4","#F8766D")) +
  ylim(0,30)

p5.supplot <- p5.sup + geom_bar(stat="identity",position=position_dodge()) + ggtitle('Supra-normal') +
    geom_hline(yintercept=5, linetype="dashed", color = "black") +
  theme(axis.text.x = element_text(angle = 90, size=14,face="bold"), 
        text = element_text(size = 14,face="bold"),
        axis.line = element_line(colour="black"),
        panel.background = element_blank(),
        plot.title = element_text(size = 16, face = "bold")) + 
    labs(y="Percentage (%)",x="Thalamic Nucleus") + 
  scale_fill_manual(breaks=c("HC","SZ"),values=c("#00BFC4","#F8766D")) + 
  ylim(0,30)

patchwork <- p5.infplot + p5.supplot + plot_layout(nrow=2) 
fig3ab <- patchwork + plot_annotation(tag_levels = 'A')
fig3ab
ggsave(file="Figure3.eps",fig3ab)
```


# Comparing deviances between groups

Calculating differences in the proportion of infra-, normal and supra-normal indivivudals in the schizophrenia and healthy control groups for each thalamic nucleus. 

```{r}
centileX <- data.frame(matrix(data=NA,nrow=18,ncol=7))
colnames(centileX) <- c("Nucleus","SZinfra","SZnorm","SZsupra","HCinfra","HCnorm","HCsupra")
centileX$Nucleus <- Nucleus
centileX$Hemisphere <- Hemisphere
group.names <- c("Schizophrenia Spectrum","Normal Control")


for (i in 1:18) {
  variablename <- paste(Nucleus[i],'_',substr(Hemisphere[i],1,1),'.',"deviance",sep="")
  centileX[i,2] <- sum(psyxsec[psyxsec$dx_group=="Schizophrenia Spectrum",variablename]==-1)
  centileX[i,3] <- sum(psyxsec[psyxsec$dx_group=="Schizophrenia Spectrum",variablename]==0)
  centileX[i,4] <- sum(psyxsec[psyxsec$dx_group=="Schizophrenia Spectrum",variablename]==1)
  centileX[i,5] <- sum(psyxsec[psyxsec$dx_group=="Normal Control",variablename]==-1)
  centileX[i,6] <- sum(psyxsec[psyxsec$dx_group=="Normal Control",variablename]==0)
  centileX[i,7] <- sum(psyxsec[psyxsec$dx_group=="Normal Control",variablename]==1)
  rm(variablename)
}

for (i in 1:18) {
  SZdata <- centileX[i,2:4]
  colnames(SZdata) <- c("infra","norm","supra")
  HCdata <- centileX[i,5:7]
  colnames(HCdata) <- c("infra","norm","supra")
  testdata <- rbind(SZdata,HCdata)
  rownames(testdata) <- c("SZ","HC")
  chisq <- chisq.test(testdata)
  #cramv <- cramersv(chisq)
  centileX[i,"Chisq"] = chisq$statistic
  centileX[i,"pval"]=chisq$p.value
  centileX[i,"cramv"]=cramersv(chisq)
}

centileX$fdr <- p.adjust(centileX[,"pval"], method="fdr")
print(centileX)
```


# Associations with Cognition

Calculating the association between SCIP composite score (cognition) and deviance scores (centiles) amd raw volumes (standardized for age, sex, race and eTIV)

```{r}
# Create function for calcuating correlations for centiles and standardized raw scores
table_rho <- function(y,Nucleus,Hemisphere, dataset = data) {
  #Setup function to calculate correlations
  corr <- data.frame(matrix(data=NA,nrow=18,ncol=6))
  colnames(corr) <- c("Nucleus","Hemisphere","centile_rho","centile_p","raw_rho","raw_p")
  corr$Nucleus <- Nucleus
  corr$Hemisphere <- Hemisphere
  
  #  corr$Nucleus <- c("AV","AV","VA","VA","VL","VL","VPL","VPL",
#                  "Pul","Pul","LGN","LGN","MGN","MGN","CM","CM","MD","MD")
#  corr$Hemisphere <- c("Right","Left","Right","Left","Right","Left","Right","Left",
#                  "Right","Left","Right","Left","Right","Left","Right","Left","Right","Left")

  for (i in 1:18) {
    centnuc <- paste("centile_",substr(corr$Hemisphere[i],1,1),'.',corr$Nucleus[i],sep="")
    resnuc <- paste("res_",substr(corr$Hemisphere[i],1,1),".",corr$Nucleus[i],sep="")
    centile = cor.test(dataset[,centnuc],dataset[,y], use = "pairwise.complete.obs",method="spearman",exact=FALSE)
    raw = cor.test(dataset[,resnuc],dataset[,y], use = "pairwise.complete.obs", method="spearman",exact=FALSE)
    corr[i,3] <- centile$estimate
    corr[i,4] <- centile$p.value
    corr[i,5] <- raw$estimate
    corr[i,6] <- raw$p.value
  }
  
  corr$centile_fdr <- p.adjust(corr[,"centile_p"], method="fdr")
  corr$raw_fdr <- p.adjust(corr[,"raw_p"], method = "fdr")
  
  return(corr)
}  
```


```{r}
draw_rho <- function(name,corr=corr_f) {
  rho.long <- melt(corr, id = c("Nucleus","Hemisphere"), 
                  measure = c("centile_rho","raw_rho"), 
                  variable.name = "Type",
                  value.name = "Rho")
  rho.long$Type <- gsub("centile_rho","Centiles",rho.long$Type)
  rho.long$Type <- gsub("raw_rho","Std Vols",rho.long$Type)

  pval.long <- melt(corr, id = c("Nucleus","Hemisphere"), 
                  measure = c("centile_fdr","raw_fdr"), 
                  variable.name = "Type",
                  value.name = "pvalue")
  label(pval.long[["pvalue"]]) <- "P-value"
  pval.long$Type <- gsub("centile_fdr","Centiles",pval.long$Type)
  pval.long$Type <- gsub("raw_fdr","Std Vols",pval.long$Type)

  corr.long <- merge(x = rho.long, y = pval.long, 
                     by.x = c("Nucleus","Hemisphere","Type"), by.y = c("Nucleus","Hemisphere","Type"))

# Heatmap for Rho 
 corr.plot <- ggplot(corr.long, aes(Nucleus, Type, fill= pvalue)) + 
   geom_tile() + 
   geom_text(aes(Nucleus,Type, label = round(Rho,3)), size=4) + 
   facet_wrap(~Hemisphere, nrow=2) +
   ggtitle(paste(name)) +
   scale_fill_gradientn(colours=c("white","yellow","orange","red"),limits=c(0,1)) +
#                       , values=rescale(c(-1,0-.Machine$double.eps,0,0+.Machine$double.eps,1)),limits=c(0,1)) +
    theme(
      panel.grid.major = element_blank(),panel.grid.minor = element_blank(), panel.background = element_blank(),
   #   axis.text = element_text(face="bold"),
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      text = element_text(size = 14,face="bold"),
      axis.text.x = element_text(angle = 90, size=14, face="bold"),
      axis.text.y = element_text(color = "black", size = 14, face = "bold"),
      plot.title = element_text(size = 16, face = "bold")
    #  axis.text.y = element_text(angle = 90, hjust = 1)
      ) +
  coord_fixed() + labs(fill="p-values")
 
  return(corr.plot)
}

```

```{r}
# Merge centile values with cognitive measures
psyxsecBEH <- merge(x = psyxsec, y = behavdata, by.x = "Subject", by.y = "XNAT.Subject.ID")

label(psyxsecBEH$Positive.Sub.Scale) <- "PANSS Positive"
label(psyxsecBEH$Negative.Sub.Scale) <- "PANSS Negative"
label(psyxsecBEH$General.Sub.Scale) <- "PANSS General"
label(psyxsecBEH$Total.Z.Score) <- "SCIP Total Score"

for (i in 1:18) {
  nucname <- paste("n",substr(Hemisphere[i],1,1),'.',Nucleus[i],sep="")
  resname <- paste("res_",substr(Hemisphere[i],1,1),'.',Nucleus[i],sep="")
  tp <- lm(eval(as.name(nucname)) ~ Age + Race + Sex + neTIV, data = psyxsecBEH)
  psyxsecBEH[,resname] <- rstandard(tp)
}

psyxsecBEH.SZ <- psyxsecBEH[psyxsecBEH$dx_group=="Schizophrenia Spectrum",]
psyxsecBEH.HC <- psyxsecBEH[psyxsecBEH$dx_group=="Normal Control",]

psyxsecBEH$Sex <- dplyr::recode(psyxsecBEH$Sex,
                                "F" = "Female",
                                "M" = "Male")
```

## Plotting correlation schores

```{r}
my.render.cont <- function(x) {
  with(stats.apply.rounding(stats.default(x), digits=4), c("",
                                                           "Mean (SD)"=sprintf("%s (&plusmn;%s)", MEAN, SD)))
}

my.render.cat <- function(x) {
  c("", sapply(stats.default(x), function(y) with(y, sprintf("%d (%0.0f %%)", FREQ, PCT))))
}

table1(~ Age + Sex + Race + Total.Z.Score | dx_group, data = psyxsecBEH, render.continuous = my.render.cont, render.categorical=my.render.cat)
```

Spearman Rho correlation with cognition using centiles and standardized (corrected for covariates) raw volumes.

```{r fig2, fig.width = 7,  fig.height=8, dev='png',dpi=300}
sz.corr = table_rho("Total.Z.Score",Nucleus,Hemisphere,dataset=psyxsecBEH.SZ)
sz.plot <- draw_rho("Schizophrenia Group",corr=sz.corr)

sz.corr
sz.plot

ggsave(file="Figure4.svg",sz.plot,width=7,height=4.5,units="in",dpi=600)
```

# Supplemental: Age distribution figures

Age distributions comparing HCP and schizophrenia datasets.

```{r}
data$Dataset <- factor(data$Dataset, levels=c("HCP","PSYXSEC"), 
                       labels=c("HCP Lifespan","Schizophrenia Cohort"))

allagehist <- ggplot(data,aes(x=Age, color=Dataset,fill=Dataset)) + 
  geom_histogram(alpha=.2,binwidth=3) +
  geom_density(aes(y=after_stat(count*3)),alpha=.25) +
  theme_classic() + 
  ylab("Number of Scans") + ylim(c(0,300)) +
  scale_fill_manual(breaks=c("HCP Lifespan","Schizophrenia Cohort"),values=c("grey","#FF0000")) +
  scale_color_manual(breaks=c("HCP Lifespan","Schizophrenia Cohort"),values=c("grey","#FF0000"))


# show on page
allagehist

# Print on page and save as eps file
ggsave("agedist_dataset.eps",allagehist,dpi=300)
```

Age distributions between diagnoses in psychosis cohort.
```{r}
psyagehist <- ggplot(psyxsec,aes(x=Age,color=dx_group,fill=dx_group)) +
  geom_histogram(alpha=.2,binwidth=2) +
  geom_density(aes(y=after_stat(count*2)),alpha=.25) +
  theme_classic() + 
  ylab("Number of Scans") + ylim(c(0,80)) +
  guides(fill=guide_legend(title="Diagnosis"),color=guide_legend(title="Diagnosis")) + 
  scale_fill_manual(breaks=c("Normal Control",
                             "Schizophrenia Spectrum"),values=c("#00BFC4","#F8766D")) +
  scale_color_manual(breaks=c("Normal Control",
                             "Schizophrenia Spectrum"),values=c("#00BFC4","#F8766D"))

# show on page
psyagehist

# save as eps file
ggsave("agedist_diagnosis.eps",psyagehist,dpi=300)
```